---
title: "COVID-19 in US"
# subtitle: (Lunch and Learn R) 
# author: Dmitry Gorodnichy
date: "on `r format(Sys.time(), '%d %B, %Y')`"
runtime: shiny
output: 
  flexdashboard::flex_dashboard:
    social: menu
    source_code: embed
    orientation: columns
    vertical_layout: fill
---
<!-- session008.Rmd = 01-board-App.Rmd -->
<!-- Run this code at https://itrack.shinyapps.io/covid/session008.Rmd -->
<!-- Learn R and how this code is developed at https://ivi-m.github.io/R-Ottawa/101.html -->

```{r readCovidUS , include=FALSE}
library(flexdashboard)

knitr::opts_chunk$set(echo = FALSE)
source("01-read.R")
dtGeo <- readGeo()
dtGeo[city == 'New York City', city:='New York']
dtUS <- readCovidUS() 
dateMax<- dtUS$date %>% max %>% ymd; dateMax
```

# Settings {.sidebar}
```{r input daysToTrack showN}
selectInput("state", "Choose State:",
            multiple=F, 
            selected = "New York",
            choices = dtUS$state %>% unique
)

sliderInput("showN", 
            label = "Narrow output to (put 0 to show all):",
            pre = "top ",
            post  = " cities",
            min = 0, max = 50, value = 10)

sliderInput("daysToTrack", label = "Days to track:",
            min = 7, max = 90, value=30)

checkboxInput("scale", "Keep scale constant", F)
```


```{r r.dt, include=FALSE}
# set my.state ----



r.dt <- reactive({
  
  my.state="New York"
  my.state=input$state 
  dt <- dtGeo [ dtUS [state == my.state], on=c("country" , "state" , "city")]; dt
  
  showN = 6
  showN <- input$showN
  if ( showN > 0 )
    dt <- dt %>% covid.reduceToTopNCitiesToday(showN);dt 
  
  dt[, recovered:=NULL]
  
  if (F) {
    dt[, confirmedGrowth := round ( confirmed / shift(confirmed,1) , 2)] [ ] # aka reproduction rate, R0
    
    dt[, confirmedPercentage:=round(confirmed/population*100,1)] []
    dt[, confirmedTotal := cumsum(confirmed)] []
    dt[, confirmedTotalPercentage := round(confirmedTotal/population*100,1)]

    dt[, deathPercentage:=round(deaths/population*100,1)][]
    dt[, deathTotal := cumsum(deaths)][]
    dt[, deathTotalPercentage := round(deathTotal/population*100,1)]
  } else {
    
    colsCases <- c("confirmed", "deaths")
    colsGrowth <- paste0(colsCases, "Growth")
    colsPercentage <- paste0(colsCases, "Percentage")
    colsTotal <- paste0(colsCases, "Total")
    colsTotalPercentage  <- paste0(colsCases, "TotalPercentage")
    
    dt[ ,  (colsGrowth) := round ( .SD / shift(.SD, 1) , 2) , .SDcols = colsCases] [ ] 
    dt[ ,  (colsPercentage) := lapply(.SD,  function(x) { round(x/population*100,1)  }   ), .SDcols = colsCases] []
    dt[ ,  (colsTotal) := cumsum(.SD), .SDcols = colsCases] []
    dt[ ,  (colsTotalPercentage) :=  lapply(.SD,  function(x) { round(x/population*100,1)  }   ), .SDcols = colsTotal] []
  }
  
  dt[, deathRate := round (deaths/confirmed*100,1)]
  dt[, deathRateAverage := round (deaths/confirmedTotal*100,1)]
  
})

```



# Map

## Column name not printed {data-width=1000}

### Map

```{r}

library(leaflet)

renderLeaflet({
  
  dtToday <- r.dt()[date==dateMax] 
  my.mean <-  dtToday$confirmed %>% mean()
  
  dtToday[, ratingcol:= ifelse(confirmed<my.mean, "yellow", "red")]
  dtToday[, ratingcol:= ifelse(confirmedGrowth<1, "yellow", "red")]
  
  
  dtToday[, strMessage:= paste0(city, 
                                ":<br>Confirmed:",
                                "<br>  Total: ", confirmedTotal, "(", confirmedTotalPercentage, "%)",
                                "<br>  Daily: ", confirmed, "(", confirmedPercentage, "% population)", 
                                "<br> Deaths:",
                                "<br>  Total: ", deathsTotal, "(", deathsTotalPercentage, "%)",
                                "<br>  Daily: ", deaths, "(", deathsPercentage, "% population)",
                                "<br>Death rate: ", 
                                "<br>  Today: ", deathRate, ". Average: ", deathRateAverage
  )  ]
  
  leaflet::leaflet(dtToday) %>% 
    addTiles() %>%
    addCircleMarkers(~lng, ~lat, 
                     color = ~ratingcol, 
                     popup = ~strMessage,
                     label = ~paste0(city, ": ", confirmed, "(", confirmedPercentage, "%) / day. R0=", confirmedGrowth) 
    ) %>% 
    addPopups(~lng,  ~lat, 
              popup = ~paste0(city, ": ", confirmed, " / day. R0=", confirmedGrowth) 
    )  %>%
    addLegend("bottomleft",
              colors = c("yellow","red"),
              labels = c(
                "Growth (R0) < 1",
                "Growth (R0) > 1"),
              opacity = 0.7)
  
})
```


# Table

## Column name not printed 


### Today  {data-height=1000}

```{r}
DT::renderDataTable({
  
  r.dt() [date==dateMax]%>% 
    DT::datatable(
      filter ="top",
      options = list(
        scrollY = TRUE,
        scrollX = TRUE)
    )
})
```


# Trends

## This caption is not printed {.tabset .tabset-fade } 

### Dynamics over time

```{r fig.width=12, fig.height=9, fig.align='center'}


r.gTrend <- reactive({
  r.dt() [ date > dateMax - input$daysToTrack] %>% 
    # covid.reduceToTopNCitiesToday(input$showN) %>% 
    ggplot() + 
    # facet_wrap( . ~ reorder(city, -confirmed), scales="free") +
    facet_wrap( . ~  reorder(city, -confirmed),
                  # ifelse(input$sortby, city, reorder(city, -confirmed)), 
                scales=ifelse(input$scale, "fixed", "free")) +  
    geom_line(aes(date, confirmed), col="orange") +
    labs(
      title= paste0("Infected per day"),
      subtitle=paste0("State: ", input$state, ". Date: ", dateMax, " (Cities are sorted by the number of infected)"),  
      caption=paste0( 
        "Data source: Johns Hopkins University U.S. Coronavirus database\n",
        "Generated by R Ottawa"
      )
    )
  
  
  
})

renderPlot( {
  print( r.gTrend() )
})

```

### Interactive plot

```{r}
library(plotly)
renderPlotly( {
  ggplotly( r.gTrend() )
})

```




# Info

<!-- ## Column -->
## Column {.tabset .tabset-fade }

### About App

#### Technical information



This App is developed as part of the *"Lunch and Learn R"* sessions, organized by Dmitry Gorodnichy within the  [R Ottawa community](https://ivi-m.github.io/R-Ottawa/). The sessions are free and open to public. They are also recorded and can be watched at YouTube. 
See  [R Ottawa - R101](https://ivi-m.github.io/R-Ottawa/101.html) portal for more details.



<!-- title: "COVID-19 in US"     -->
<!-- subtitle: "Situational Analysis Report"     -->
<!-- author: "Dmitry Gorodnichy"    -->
<!-- <!-- date: "Last updated: `r format (Sys.time(), '%d %B, %Y')`"    --> -->
<!-- date: "Last updated: `3 June 2020`"    -->



*Name*: COVID-19 Tracker  (US - Light Version)    
*Type*: Web Application. Compatible with mobile devices.    
*URL*: https://itrack.shinyapps.io/covid/us-session008.Rmd    
*License*:  [CC BY-NC](https://creativecommons.org/licenses/by-nc/4.0/legalcode).   
*Developer*: [Dmitry Gorodnichy](https://www.linkedin.com/in/dmitry-gorodnichy/)  and  [R Ottawa community](https://ivi-m.github.io/R-Ottawa/).  


*Last updated*: 3 June 2020
*Version*: 0.0.8

*Questions/Comments*: [dg@ivim.ca](email:dg@ivim.ca)
<!-- *Contact*: [dg@ivim.ca](email:dg@ivim.ca) -->



#### Disclaimer 

This work is not to be associated with any  work done by the Government of Canada. The authors are not liable for any errors that might be present in  data or data interpretation. For all official information related to Covid in Canada please visit the official [Coronavirus disease (COVID-19) site](https://www.canada.ca/en/public-health/services/diseases/coronavirus-disease-covid-19.html?utm_campaign=not-applicable&utm_medium=vanity-url&utm_source=canada-ca_coronavirus).



#### Acknowledgements

The App is developed using R in RStudio. 
The following R libraries are used:    

- Data management: data.table, dtplyr,  stringr, readxl, lubridate, magrittr.   
- Data visualization: ggplot2, plotly, leaflet, DT.  
<!-- - Machine learning: caret, GMDH.   (TBA) -->
- Publishing: shiny, rsconnect,  knitr, rmarkdown, flexdashboard.   


### Data source

This report is generated automatically. 
The COVID-19 data is taken from:   

- [Johns Hopkins University U.S. Coronavirus database](https://coronavirus.jhu.edu/us-map), which is available at this [GitHub repository](https://github.com/CSSEGISandData/COVID-19).

This database was last updated on `r dateMax`.

